---
title: "Phage and Plasmid Ecology Analyses (DWW)"
subtitle: "Day 3, Exercise 2 — UCD Bioinformatics Course"
author: "Mery Touceda-Suarez"
date: "June 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview and Steps Guide

In this exercise you will explore the ecology of phages (viruses) and plasmids recovered from wastewater metagenomes. You will work through the same workflow for both groups: prepare the data, calculate alpha diversity, and then examine community composition (beta diversity).

Use the list below as your roadmap. Each step corresponds to a section further down in this document.

1. **Load libraries** — load `tidyverse` and `vegan`, the two packages we'll use throughout.
2. **Load and prepare the data** — read in the metadata, vOTU (viral) count table, and pOTU (plasmid) count table; reshape and align them so every table has samples in the same order.
3. **Sanity-check sample order** — confirm metadata and count tables match before doing any analysis.
4. **Alpha diversity** — rarefy the count tables and calculate richness and Shannon diversity for both phages and plasmids; visualize differences across sites.
5. **Normalize abundances** — calculate RPKM (for viral/plasmid load) and TPM (for composition comparisons) using contig/sequence length and sequencing depth.
6. **Community composition (beta diversity)** — compute Bray-Curtis dissimilarities, run NMDS ordinations, test for effects of site and collection date with PERMANOVA (`adonis2`), and plot ordinations for both phages and plasmids.
7. **Discuss the open questions** — at the end of the script, use what you've learned to think through the follow-up questions about diversity, composition, and length/abundance relationships.

Work through each code chunk in order, and don't forget to update the file paths to match your own project directory before running.

---

# 1. Load Libraries

```{r libraries}
library(tidyverse)
library(vegan)
```

# 2. Load and Prepare Data

## 2.1 Metadata

```{r load-metadata}
md <- read.csv("~/Documents/GitHub/ucd_bioinformaticscourse_2026/data/metadata_day3.csv")

md <- md %>%
  arrange(desc(sample)) %>%
  # make site variable
  mutate(site = paste(country, city, plant)) %>%
  # make date character into date class
  mutate(collection_date_n = as.Date(collection_date))
```

## 2.2 Phages (vOTUs)

```{r load-votus}
votus_count <- read.table("~/Documents/GitHub/ucd_bioinformaticscourse_2026/data/votus_count_table.txt", 
                           sep = "\t", header = TRUE)

# prepare data
votus_clean <- votus_count %>%
  pivot_longer(!Contig, names_to = "sample", values_to = "counts") %>%
  mutate(sample = str_remove(sample, ".Read.Count")) %>%
  filter(sample %in% md$sample) %>%
  pivot_wider(names_from = "Contig", values_from = "counts") %>%
  arrange(desc(sample))
```

## 2.3 Plasmids (pOTUs)

```{r load-potus}
potus_count <- read.table("~/Documents/GitHub/ucd_bioinformaticscourse_2026/data/potus_count_table.txt", 
                           sep = "\t", header = TRUE)

# prepare data
potus_clean <- potus_count %>%
  pivot_longer(!Contig, names_to = "sample", values_to = "counts") %>%
  mutate(sample = str_remove(sample, ".Read.Count")) %>%
  filter(sample %in% md$sample) %>%
  pivot_wider(names_from = "Contig", values_from = "counts") %>%
  arrange(desc(sample))
```

# 3. Sanity Check: Sample Order

Before doing any analysis, always confirm that your metadata and count tables are aligned in the same sample order.

```{r check-order}
md$sample == votus_clean$sample
md$sample == potus_clean$sample
```

# 4. Alpha Diversity

## 4.1 Prepare Count Tables

```{r prep-rich}
votus_for_rich <- votus_clean %>%
  column_to_rownames(var = "sample") %>%
  mutate(across(where(is.character), as.numeric))

potus_for_rich <- potus_clean %>%
  column_to_rownames(var = "sample") %>%
  mutate(across(where(is.character), as.numeric))
```

## 4.2 Check Counts Distribution

```{r check-distribution}
summary(rowSums(votus_for_rich))
hist(rowSums(votus_for_rich))

summary(rowSums(potus_for_rich))
hist(rowSums(potus_for_rich))
```

## 4.3 Calculate Diversity Measures

```{r diversity-measures}
md$viral_rich <- specnumber(rrarefy(votus_for_rich, sample = 71456))
md$viral_shannon <- diversity(rrarefy(votus_for_rich, sample = 71456), index = "shannon")
md$plasmid_rich <- specnumber(rrarefy(potus_for_rich, sample = 63508))
md$plasmid_shannon <- diversity(rrarefy(potus_for_rich, sample = 63508), index = "shannon")
```

## 4.4 Differences per Site — Phages

```{r viral-shannon-plot}
ggplot(md, aes(x = site, y = viral_shannon)) +
  geom_boxplot(aes(color = site)) +
  xlab("") +
  ylab("Number of vOTUs") +
  theme_classic() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# ggsave("CHANGE THIS PATH! /figures/viral_shannon_site.pdf", width = 5, height = 7, units = "in")
```

## 4.5 Differences per Site — Plasmids

```{r plasmid-shannon-plot}
ggplot(md, aes(x = site, y = plasmid_shannon)) +
  geom_boxplot(aes(color = site)) +
  xlab("") +
  ylab("Number of pOTUs") +
  theme_classic() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

# 5. Community Composition

## 5.1 RPKM/TPM Abundance Normalization — Phages

```{r normalize-phages}
total_reads <- md %>%
  select(sample, total_reads)

# load votus length
votus_length <- read_csv("~/Documents/GitHub/ucd_bioinformaticscourse_2026/data/phages_summary.csv") %>%
  select(c("votu", "contig_length"))

# convert count table to long format
votus_count_long <- votus_clean %>%
  pivot_longer(cols = -c("sample"), values_to = "count", names_to = "votu")

votus_count_long_plusreads <- left_join(votus_count_long, total_reads, by = "sample")
votus_count_long_plusreads_pluslength <- left_join(votus_count_long_plusreads, votus_length, by = "votu")

# RPKM (for viral load)
rpkm_votu_count <- votus_count_long_plusreads_pluslength %>%
  mutate(RPKM = (count * 1e6) / (contig_length * total_reads)) %>%
  dplyr::select(-c("count", "total_reads", "contig_length")) %>%
  pivot_wider(names_from = "sample", values_from = "RPKM") %>%
  column_to_rownames(var = "votu") %>%
  drop_na()

# TPM (for composition comparisons, beta)
tpm_votu_count <- votus_count_long_plusreads_pluslength %>%
  mutate(RPK = (count * 1e6) / contig_length) %>%
  group_by(sample) %>%
  mutate(TPM = RPK / sum(RPK, na.rm = TRUE) * 1e6) %>%
  ungroup() %>%
  dplyr::select(-c("count", "total_reads", "contig_length", "RPK")) %>%
  pivot_wider(names_from = "sample", values_from = "TPM") %>%
  column_to_rownames(var = "votu") %>%
  drop_na()
```

## 5.2 RPKM Abundance Normalization — Plasmids

```{r normalize-plasmids}
potus_length <- read_csv("~/Documents/GitHub/ucd_bioinformaticscourse_2026/data/plasmids_summary.csv") %>%   # change for your selection
  select(c("seq_name", "length")) %>%
  rename(potu = seq_name)

potus_count_long <- potus_clean %>%
  pivot_longer(cols = -c("sample"), values_to = "count", names_to = "potu")

potus_count_long_plusreads <- left_join(potus_count_long, total_reads, by = "sample")
potus_count_long_plusreads_pluslength <- left_join(potus_count_long_plusreads, potus_length, by = "potu")

rpkm_potu_count <- potus_count_long_plusreads_pluslength %>%
  mutate(RPKM = (count * 1e9) / (length * total_reads)) %>%
  dplyr::select(-c("count", "total_reads", "length")) %>%
  pivot_wider(names_from = "sample", values_from = "RPKM") %>%
  column_to_rownames(var = "potu") %>%
  drop_na()
```

## 5.3 Community Composition — Phages

```{r phage-composition}
viral.bray <- vegdist(t(rpkm_votu_count), method = "bray")

# ordination (non-metric multidimensional scaling)
viral.nmds <- metaMDS(viral.bray)
md$bracken.Axis01 <- viral.nmds$points[, 1]
md$bracken.Axis02 <- viral.nmds$points[, 2]
viral.nmds$stress # 0.1588582

# stats
adonis2(viral.bray ~ site, data = md, permutations = 999, method = "bray")
```

### Ordination Plot — Phages

```{r phage-ordination-plot}
ggplot(md, aes(bracken.Axis01, bracken.Axis02)) +
  geom_point(aes(alpha = collection_date_n, color = site), size = 3.5) +
  stat_ellipse(aes(color = site)) +
  theme_classic() +
  theme(legend.position = "right", text = element_text(size = 12))

# ggsave("CHANGE THIS PATH! /viral_composition_bydate.pdf", device = "pdf", width = 9, height = 5, units = "in")
```

## 5.4 Community Composition — Plasmids

```{r plasmid-composition}
plasmid.bray <- vegdist(t(rpkm_potu_count), method = "bray")

# ordination (non-metric multidimensional scaling)
plasmid.nmds <- metaMDS(plasmid.bray)
md$bracken.Axis01 <- plasmid.nmds$points[, 1]
md$bracken.Axis02 <- plasmid.nmds$points[, 2]
plasmid.nmds$stress # 0.1588582

# stats
adonis2(plasmid.bray ~ site, data = md, permutations = 999, method = "bray")
```

### Ordination Plot — Plasmids

```{r plasmid-ordination-plot}
ggplot(md, aes(bracken.Axis01, bracken.Axis02)) +
  geom_point(aes(alpha = collection_date_n, color = site), size = 3.5) +
  stat_ellipse(aes(color = site)) +
  theme_classic() +
  theme(legend.position = "right", text = element_text(size = 12))

# ggsave("CHANGE THIS PATH!/plasmid_composition_bydate.pdf", device = "pdf", width = 9, height = 5, units = "in")
```

# 6. Extra Questions

Now that you've completed the workflow, you can try to answer the following questions:

1. What variables are determining viral and phage diversity and community composition, apart from the location and the time?
2. Do plasmid and viral diversity correlate?
3. Do their community compositions vary together?
4. What are the most abundant phages/plasmids? What is their length distribution?
5. Do we see changes in phage/plasmid abundance associated with their length? (i.e. are shorter viruses more abundant?)
